\pkg{shorts}: An R Package for Modeling Short Sprints

### 2020-12-30 paper.Rmd
Preprint version

---
documentclass: jss
author:
  - name: Mladen Jovanović
    affiliation: |
      | University of Belgrade
    address: |
      | Faculty of Sport and Physical Education
      | Belgrade, Serbia
    email: \email{coach.mladen.jovanovic@gmail.com}
  - name: Jason D. Vescovi
    affiliation: |
      | University of Toronto
    address: |
      | Faculty of Kinesiology and Physical Education
      | Graduate School of Exercise Science
      | Toronto, ON Canada
    email: \email{vescovij@gmail.com}
title:
  formatted: "\\pkg{shorts}: An R Package for Modeling Short Sprints"
abstract: >
  Short sprint performance is one of the most distinguishable and admired physical trait in sports. Short sprints have been modeled using the mono-exponential equation that involves two parameters: (1) maximum sprinting speed (MSS) and (2) relative acceleration (TAU). The most common methods to assess short sprint performance are with a radar gun or timing gates. In this paper, we: 1) provide the \pkg{shorts} package that can model sprint timing data from these two sources; 2) discuss potential issues with assessing sprint time (synchronization and flying start, respectively); and 3) provide model definitions within the \pkg{shorts} package to help alleviate errors within the subsequent parameter outcomes.
keywords:
  formatted: [sprinting, sports, modeling]
preamble: >
  \usepackage{amsmath}
output:
  bookdown::pdf_book:
    base_format: rticles::jss_article
citation_package: natbib
bibliography: [references.bib, packages.bib]
link-citations: yes
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
require(knitr)
require(shorts)
require(tidyverse)
require(bookdown)

@@

options("width" = 70)

knitr::opts_chunk$set(
  comment = "#>",
  collapse = TRUE,
  cache = FALSE,
  fig.retina = 0.8, # figures are either vectors or 300 dpi diagrams
  dpi = 300,
  out.width = "100%",
  fig.align = "center",
  fig.width = 6,
  fig.height = 6 * 0.618, # 1 / phi
  fig.show = "hold",
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  width = 70
)

@@

# Set rounding
op <- options()
options(digits = 3)
```

@@

Short sprint performance is one of the most distinguishable and admired physical trait in sports. Short sprints, commonly performed in most team sports (e.g., soccer, field hockey, handball, football, etc.), are defined as maximal running from a stand still position over a distance that doesn't result in deceleration at the end. Peak anaerobic power is achieved within the first few seconds (<5s) of maximal efforts [@mangineSpeedForcePower2014], whereas the ability to achieve maximal sprint speed varies based on the type of sport. For example, track and field sprinters are trained to achieve maximal speed later in a race (i.e., 50-60m) [@ward-smithEnergyConversionStrategies2001], but team sport athletes have sport-specific attributes and reach it much sooner (i.e., 30-40m)[@brownAssessmentLinearSprinting2004]. Regardless of the differences in kinematics between athletes, evaluating short sprint performance is routinely included within a battery of fitness tests for a wide range of sports

The use of force plates is considered the gold standard for assessing mechanical properties of sprinting; however, there are logistical and financial challenges to capturing the profile of an entire sprint [@morinSimpleMethodComputing2019; @samozinoSimpleMethodMeasuring2016]. Radar and laser technology are frequently used laboratory-grade methods [@buchheitMechanicalDeterminantsAcceleration2014; @edwardsSprintAccelerationCharacteristics2020; @jimenez-reyesRelationshipVerticalHorizontal2018; @marcote-pequenoAssociationForceVelocity2019] but not normally accessible to practitioners working in sports. Undoubtedly, the most common method available and used to evaluate sprint performance are timing gates. Often multiple gates are positioned at varying distances to capture split times (e.g., 5, 10, 20m), which can now be incorporated into the method for determining sprint mechanical properties [@morinSimpleMethodComputing2019; @samozinoSimpleMethodMeasuring2016]. This approach presents an advantage to practitioners who can use the outcomes to describe individual differences, quantify the effects of training interventions, and better understanding the limiting factors of performance. The \pkg{shorts} package [@R-shorts], written in the R language [@R-base], represents an open-source tool to help sport scientists translate raw timing data into detailed mechanical outcomes through mathematical modeling [@morinSimpleMethodComputing2019; @samozinoSimpleMethodMeasuring2016].

@@

knitr::kable(
  athletes,
  caption = "Four athletes with different MSS and MAC parameters.",
  digits = 2
)
```

Figure \@ref(fig:four-athletes-kinematics) depicts distance, velocity, and acceleration over time (from 0 to 6s).

```{r four-athletes-kinematics, echo=FALSE, fig.cap="Kinematic characteristic of four athletes with different MSS and MAC parameters over a period of 0 to 6seconds."}
kinematics <- expand_grid(
  athletes,
  time = seq(0, 6, length.out = 1000)
) %>%
  mutate(
    distance = predict_distance_at_time(time, MSS, TAU),
    velocity = predict_velocity_at_time(time, MSS, TAU),
    acceleration = predict_acceleration_at_time(time, MSS, TAU)
  )

@@

Plotting acceleration against velocity (Figure \@ref(fig:four-athletes-profile)), we will get Acceleration-Velocity profile, which is linear, according to the mathematical model.

@@

Short sprints profiling is usually performed by: (1) measuring split times using timing gates (i.e., positioned at various distances, e.g., 5, 10, 20, 30, 40m), (2) getting a velocity trace using a radar gun. Estimation of MSS and TAU parameters from equation \@ref(eq:velocity-time) is performed in \pkg{shorts} package using non-linear least squares regression implemented in the `nls` function in the \pkg{base} R [@R-base] and `nlme` function in the \pkg{nlme} package [@R-nlme] for the mixed-effect models.

@@

Let's consider an example of an athlete with MSS equal to 9 $ms^-1$, TAU equal to 1.3, and MAC equal to 6.92 $ms^-2$ performing 40m sprint with timing gates positioned at each 10m split. For split times, distance is a predictor, and time is the outcome variable, thus the equation \@ref(eq:velocity-time) becomes:

@@

$W$ in equation \@ref(eq:time-distance) represents Lambert's W function [@R-LambertW]. MSS and TAU parameters are estimated using `model_using_splits` function:

@@

Maximal relative power (PMAX) from the output is estimated using $\frac{MSS \times MAC}{4}$, which disregards the air resistance. `time_correction` and `distance_corection` parameters will be covered later in the article.

Besides providing *residual standard error* (RSE), \pkg{shorts} functions provide additional model fit estimators. Additional information can be gained by exploring the returned object, particularly object returned from the `nls` function:

```{r}
summary(m1$model)
```

Once we have estimated MSS and TAU, we can use `predict_` family of functions to predict various relationships (i.e. time at distance, acceleration at distance, velocity at time):

@@

To estimate force production at distance or time (using `predict_force_at_distance` and `predict_force_at_time` functions), as well as power production (using `predict_power_at_distance` and `predict_power_at_time` functions), one needs to take into account the air resistance. Air resistance (N) is estimated using `get_air_resistance` function, which takes velocity, body mass (kg), body height (m), barometric pressure (Torr), air temperature ($C^\circ$), and wind velocity ($ms^-1$) as parameters (please refer to @arsacModelingEnergetics100m2002, @samozinoSimpleMethodMeasuring2016, and @vaningenschenauCanCyclePower1991 for more information):

@@

When estimating force and power, the air resistance parameters can be set using `"..."`, which are forwarded to the `get_air_resistance`:

@@

The easiest way to get all kinematics for short sprints is to use `predict_kinematics` functions:

@@

df <- pivot_longer(data = df, cols = -2)

@@

@@

Another valuable addition for sport scientists and coaches is the ability to determine the distances and times where 90% of maximum sprinting speed is reached, or where peak power is within 90% range. To identify these values, \pkg{shorts} package comes with `find_` family of functions:

@@

Sprint performance is often evaluated with a group of athletes (e.g., soccer club) representing a single strata of interest. Sports scientists can estimate individual profiles, or utilize mixed-effects models. To perform mixed-effects models in \pkg{shorts} for split times, one can use `mixed_model_using_splits` function. To demonstrate this functionality, we load the `split_times` dataset provided in the \pkg{shorts} package:

@@

Additional information about mixed-effects model performed using the `nlme` package [@R-nlme] can be obtained using `summary`:

@@

@@

df <- pivot_longer(df, cols = c(-1, -3))

@@

Let's consider the same example of an athlete with MSS equal to 9$ms^-1$, TAU equal to 1.3, and MAC equal to 6.92$ms^-2$ performing 40m sprint with velocity estimated using radar run (in this case with 1Hz sampling rate).

@@

@@

Mixed-effects model using radar data is done using `mixed_model_using_radar` function. To perform mixed model, let's load data that comes with \pkg{shorts} package.

@@

@@

One source of error in the modeled estimation using a radar gun is the time synchronization. In theory, synchronization is ideal when a sprint is initiated at $t=0$ (i.e., $v(t=0) = 0$). In practice, this is often not the case. Let's use our athlete and add and deduct 0.5s to simulate an error in synchronization and its effect on estimated MSS and TAU.

@@

As can be seen from the example, TAU is affected by an error in synchronization of time with velocity. The potential solution incorporated into the \pkg{shorts} package involves estimation of the *time correction* parameter using the following equation:

@@

This model is incorporated in the `model_using_radar_with_time_correction` function:

@@

When using `predict_` family of functions, one can provide estimated time correction to get predictions at original time scale.

@@

When it comes to mixed-model approach, time correction can be modeled as a fixed effect or random effect using the `mixed_model_using_radar_with_time_correction` function.

@@

Let's imagine we have two twin brothers with same short sprint characteristics: MSS equal to 9$ms^-1$, TAU equal to 1.3, and MAC equal to 6.92$ms^-2$. Let's call them John and Jack. They both perform 40m sprint using timing gates set at 5, 10, 20, 30, and 40m. The initial timing gate at the start (i.e., $d=0m$) serves to activate the timing system (i.e., when they cross the beam).

John represents the *theoretical model*, in which we assume that the initial force production and the timing initiation are perfectly synchronized. Jack, on the other hand, represents a *practical model*, and decides to move slightly behind the initial timing gate (i.e. for 0.5m) and use body rocking to initiate the sprint start. In other words, Jack is using a *flying start*, a common scenario when testing field sports athletes.  Let's see how their sprint outcomes differ.

@@

sprint_parameters <- rbind(
  unlist(john_profile$parameters),
  unlist(jack_profile$parameters)
)

@@

Below is a simulation sprint with 5, 10, 20, 30, 40, and 50m splits, with MSS and MAC varying from 6 to 9 ($ms^-1$ and $ms^-2$ respectively), and flying start distance varying from 0 to 1m.

@@

@@

  params <- data.frame(t(unlist(model$parameters)))

@@

The following image demonstrates the effect of flying start distance on estimated MSS:

@@

As can be seen from the image, MSS is underestimated as flying start distance increases. The following image demonstrates the effect of flying start distance on estimated MAC:

@@

Model residuals are also affected by flying start distance. The shape of residuals distribution depends on number and splits utilized (e.g., 10, 20, 30, 40m versus 5, 15, 30m), but here we can see the effect of the flying start distance on the model residuals per split distance utilized in our simulation:

@@

ggplot(
  model_df,
  aes(y = residuals, x = distance, color = flying_start_distance, group = group)
) +
  theme_bw(8) +
  geom_line(alpha = 0.3) +
  facet_grid(MSS_string ~ MAC_string) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  scale_color_gradientn(colours = terrain.colors(5, rev = FALSE)) +
  xlab("Distance (m)") +
  ylab("Predicted time - observed time (s)") +
  theme(legend.position = "top") + 
  labs(color = "Flying start distance")
```

If we merge individual facets (i.e., combinations of MSS and MAC), we can get simpler images conveying issues with residuals when there is a flying start:

```{r}
ggplot(
  model_df,
  aes(y = residuals, x = distance, color = flying_start_distance, group = group)
) +
  theme_bw(8) +
  geom_line(alpha = 0.3) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  scale_color_gradientn(colours = terrain.colors(5, rev = FALSE)) +
  xlab("Distance (m)") +
  ylab("Predicted time - observed time (s)") +
  theme(legend.position = "top") + 
  labs(color = "Flying start distance")
```

@@

A potential solution is to use a correction factor - the recommendation in the literature is +0.5sec [@haugenSprintMechanicalProperties2020; @haugenSprintMechanicalVariables2019]. Interestingly, the average difference between using timing gates and a block start for 40 m sprint time was 0.27 sec [@haugenDifferenceStartImpact2012]. So, while a timing correction factor is warranted to avoid subsequent errors in estimates of kinetic variables (e.g., overestimate power), a correction factor that is too large will have the opposite effect (e.g., underestimate power).

When implementing *time correction*, equation \@ref(eq:time-distance) becomes:

@@

Rather than providing *apriori* time correction from the literature, \pkg{shorts} package provides an estimation of this parameter from the data provided, together with MSS and TAU. To estimate time correction, we use `model_using_splits_with_time_correction` function. Here is how we can estimate Jack parameters using either provided time correction (e.g., +0.3 and +0.5s) or estimated time correction:

@@

jack_parameters <- rbind(
  unlist(john_profile$parameters),
  unlist(jack_profile$parameters),
  unlist(jack_profile_fixed_time_short$parameters),
  unlist(jack_profile_fixed_time_long$parameters),
  unlist(jack_profile_time_estimated$parameters)
)

@@

In Jack's case, both +0.3s fixed time correction and time correction estimation yield parameters closer to John's (i.e. true parameters).

Another model definition, which is a novel approach implemented in the \pkg{shorts} packages, is to utilize *distance correction*, besides time correction. Thus, equation \@ref(eq:time-distance) becomes:

@@

This model is implemented in `model_using_splits_with_corrections` function. Below are the model estimates:

@@

jack_parameters <- rbind(
  unlist(john_profile$parameters),
  unlist(jack_profile$parameters),
  unlist(jack_profile_fixed_time_short$parameters),
  unlist(jack_profile_fixed_time_long$parameters),
  unlist(jack_profile_time_estimated$parameters),
  unlist(jack_profile_distance_correction$parameters)
)

@@

As can be seen from the results, adding distance correction results in correctly estimating Jack's sprint parameters. There are a few issues with this model definition. Besides being novel and still not validated with actual data, distance correction model has four parameters to estimate, which implies that at least five sprint splits are needed. This imposes practical limitations, since acquiring six timing gate (one for the start and five for splits) might be practically troublesome.

@@

  params <- rbind(
    data.frame(
      model = "No correction",
      t(unlist(no_correction$parameters))
    ),
    data.frame(
      model = "Fixed correction +0.3s",
      t(unlist(fixed_correction_short$parameters))
    ),
    data.frame(
      model = "Fixed correction +0.5s",
      t(unlist(fixed_correction_long$parameters))
    ),
    data.frame(
      model = "Time correction",
      t(unlist(time_correction$parameters))
    ),
    data.frame(
      model = "Time and distance correction",
      t(unlist(time_dist_correction$parameters))
    )
  )

@@

ggplot(
  model_df,
  aes(y = residuals, x = distance, color = flying_start_distance, group = group)
) +
  theme_bw(8) +
  geom_line(alpha = 0.3) +
  facet_wrap(~model) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  scale_color_gradientn(colours = terrain.colors(5, rev = FALSE)) +
  xlab("Distance (m)") +
  ylab("Predicted time - observed time (s)") +
  theme(legend.position = "top") + 
  labs(color = "Flying start distance")
```

@@

Time correction and time and distance corrections are also implemented in the mixed-models using `mixed_model_using_splits_with_time_correction` and `mixed_model_using_splits_with_corrections`. We will showcase their use at the end of this article.

@@

Starting behind the initial timing gate represent only one issue (i.e. flying start). In this section, we simulate one more issue to check the sensitivity of the presented models to other (less common) perturbations when performing field testing.

>>>>

---
documentclass: jss
author:
  - name: Mladen Jovanović
    affiliation: |
      | University of Belgrade
    address: |
      | Faculty of Sport and Physical Education
      | Belgrade, Serbia
    email: \email{coach.mladen.jovanovic@gmail.com}
  - name: Jason D. Vescovi
    affiliation: |
      | University of Toronto
    address: |
      | Faculty of Kinesiology and Physical Education
      | Graduate School of Exercise Science
      | Toronto, ON Canada
    email: \email{vescovij@gmail.com}
title:
  formatted: "\\pkg{shorts}: An R Package for Modeling Short Sprints"
DOI: "10.31236/osf.io/4jw62"
URL: https://osf.io/preprints/sportrxiv/4jw62/
abstract: >
  Short sprint performance is one of the most distinguishable and admired physical trait in sports. Short sprints have been modeled using the mono-exponential equation that involves two parameters: (1) maximum sprinting speed (MSS) and (2) relative acceleration (TAU). The most common methods to assess short sprint performance are with a radar gun or timing gates. In this paper, we: 1) provide the \pkg{shorts} package that can model sprint timing data from these two sources; 2) discuss potential issues with assessing sprint time (synchronization and flying start, respectively); and 3) provide model definitions within the \pkg{shorts} package to help alleviate errors within the subsequent parameter outcomes.
keywords:
  formatted: [sprinting, sports, modeling]
preamble: >
  \usepackage{amsmath}
output:
  bookdown::pdf_book:
    base_format: rticles::jss_article
    extra_dependencies: ["flafter"]
citation_package: natbib
bibliography: [references.bib, packages.bib]
link-citations: yes
header-includes: >
  \usepackage[htt]{hyphenat}
  \usepackage{booktabs}
  \usepackage{microtype}
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
require(knitr)
require(shorts)
require(tidyverse)
require(bookdown)
require(magick)

pdf2png <- function(path) {
  # only do the conversion for non-LaTeX output
  if (knitr::is_latex_output()) 
    return(path)
  path2 <- xfun::with_ext(path, "png")
  img <- magick::image_read_pdf(path)
  magick::image_write(img, path2, format = "png")
  path2
}

@@

knitr::opts_chunk$set(
  comment = "#>",
  collapse = TRUE,
  cache = FALSE,
  fig.retina = 0.8, # figures are either vectors or 300 dpi diagrams
  dpi = 600,
  dev = "png",
  fig.process = pdf2png,
  out.width = "100%",
  fig.align = "center",
  fig.width = 6,
  fig.height = 6 * 0.618, # 1 / phi
  fig.show = "hold",
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  width = 70
)

@@

# Set rounding
op <- options()

options(
  digits = 3,
  "width" = 70,
  scipen = 999)
```

@@

Short sprint performance is one of the most distinguishable and admired physical trait in sports. Short sprints, commonly performed in most team sports (e.g., soccer, field hockey, handball, football, etc.), are defined as maximal running from a stand still position over a distance that doesn't result in deceleration at the end. Peak anaerobic power is achieved within the first few seconds (<5 s) of maximal efforts [@mangineSpeedForcePower2014], whereas the ability to achieve maximal sprint speed varies based on the type of sport. For example, track and field sprinters are trained to achieve maximal speed later in a race (i.e., 50-60 m) [@ward-smithEnergyConversionStrategies2001], but team sport athletes have sport-specific attributes and reach it much sooner (i.e., 30-40 m)[@brownAssessmentLinearSprinting2004]. Regardless of the differences in kinematics between athletes, evaluating short sprint performance is routinely included within a battery of fitness tests for a wide range of sports

The use of force plates is considered the gold standard for assessing mechanical properties of sprinting; however, there are logistical and financial challenges to capturing the profile of an entire sprint [@morinSimpleMethodComputing2019; @samozinoSimpleMethodMeasuring2016]. Radar and laser technology are frequently used laboratory-grade methods [@buchheitMechanicalDeterminantsAcceleration2014; @edwardsSprintAccelerationCharacteristics2020; @jimenez-reyesRelationshipVerticalHorizontal2018; @marcote-pequenoAssociationForceVelocity2019] but not normally accessible to practitioners working in sports. Undoubtedly, the most common method available and used to evaluate sprint performance are timing gates. Often multiple gates are positioned at varying distances to capture split times (e.g., 5, 10, 20 m), which can now be incorporated into the method for determining sprint mechanical properties [@morinSimpleMethodComputing2019; @samozinoSimpleMethodMeasuring2016]. This approach presents an advantage to practitioners who can use the outcomes to describe individual differences, quantify the effects of training interventions, and better understanding the limiting factors of performance. The \pkg{shorts} package [@R-shorts], written in the R language [@R-base], represents an open-source tool to help sport scientists translate raw timing data into detailed mechanical outcomes through mathematical modeling [@morinSimpleMethodComputing2019; @samozinoSimpleMethodMeasuring2016].

@@

knitr::kable(
  athletes,
  caption = "Four athletes with different MSS and MAC parameters.",
  digits = 2,
  booktabs = TRUE
)
```

Figure \@ref(fig:four-athletes-kinematics) depicts distance, velocity, and acceleration over time (from 0 to 6 s).

```{r four-athletes-kinematics, echo=FALSE, fig.cap="Kinematic characteristic of four athletes with different MSS and MAC parameters over a period of 0 to 6 seconds."}
kinematics <- expand_grid(
  athletes,
  time = seq(0, 6, length.out = 1000)
) %>%
  mutate(
    distance = predict_distance_at_time(time, MSS, TAU),
    velocity = predict_velocity_at_time(time, MSS, TAU),
    acceleration = predict_acceleration_at_time(time, MSS, TAU)
  )

@@

Plotting acceleration against velocity (Figure \@ref(fig:four-athletes-profile)), we will get *Acceleration-Velocity Profile*, which is linear, according to the mathematical model. If the athlete's body mass (kg) is known, as well as additional air resistance parameters (see [Air resistance and the calculation of force and mechanical power] section of this paper), *Force-Velocity Profile* can be estimated (see [Force-Velocity profile] section of this paper).

@@

Short sprints profiling is usually performed by: (1) measuring split times using timing gates (i.e., positioned at various distances, e.g., 5, 10, 20, 30, 40 m), (2) getting a velocity trace using a radar gun. Estimation of MSS and TAU parameters from equation \@ref(eq:velocity-time) is performed in \pkg{shorts} package using non-linear least squares regression implemented in the `nls()` function in the \pkg{base} R [@R-base] and `nlme()` function in the \pkg{nlme} package [@R-nlme] for the mixed-effect models.

@@

Let's consider an example of an athlete with MSS equal to 9 $ms^{-1}$, TAU equal to 1.3, and MAC equal to 6.92 $ms^{-2}$ performing 40m sprint with timing gates positioned at each 10m split. For split times, distance is a predictor, and time is the outcome variable, thus the equation \@ref(eq:velocity-time) becomes:

@@

$W$ in equation \@ref(eq:time-distance) represents Lambert's W function [@R-LambertW]. MSS and TAU parameters are estimated using `model_using_splits()` function:

@@

Maximal relative power (PMAX) from the output is estimated using $\frac{MSS \times MAC}{4}$, which disregards the air resistance. `time_correction` and `distance_corection` parameters will be covered later in the paper.

Besides providing *residual standard error* (RSE), \pkg{shorts} functions provide additional model fit estimators. Additional information can be gained by exploring the returned object, particularly object returned from the `nls()` function:

```{r}
summary(m1)
```

To extract estimated model parameters, use S3 `coef()` method:

```{r}
coef(m1)
```

To create a simple plot of the model, use S3 `plot()` method, which returns \pkg{ggplot2} [@R-ggplot2] object:

```{r}
plot(m1) + theme_bw(8)
```

Once we have estimated MSS and TAU, we can use `predict_XXX()` family of functions to predict various relationships (i.e., time at distance, acceleration at distance, velocity at time, etc.):

@@

To estimate force production at distance or time (using `predict_force_at_distance()` and `predict_force_at_time()` functions), as well as power production (using `predict_power_at_distance()` and `predict_power_at_time()` functions), one needs to take into account the air resistance. Air resistance (N) is estimated using `get_air_resistance()` function, which takes velocity, body mass (kg), body height (m), barometric pressure (Torr), air temperature ($C^\circ$), and wind velocity ($ms^-1$) as parameters (please refer to @arsacModelingEnergetics100m2002, @samozinoSimpleMethodMeasuring2016, and @vaningenschenauCanCyclePower1991 for more information):

@@

When estimating force and power, the air resistance parameters can be set using `"..."`, which are forwarded to the `get_air_resistance()`:

@@

The easiest way to get all kinematics and kinetics for short sprints is to use `predict_kinematics()` function:

@@

variable_names <- colnames(df)

df <- pivot_longer(data = df, cols = -2) %>%
  mutate(name = factor(name, levels = variable_names))

@@

These kinematic and kinetic variables are utilized in [Force-Velocity profile] estimation, which is covered later in this paper.

@@

Another valuable addition for sport scientists and coaches is the ability to determine the distances and times where 90% of maximum sprinting speed is reached, or where peak power is within 90% range. To identify these values, \pkg{shorts} package comes with `find_XXX()` family of functions:

@@

Sprint performance is often evaluated with a group of athletes (e.g., soccer club) representing a single strata of interest. Sports scientists can estimate individual profiles, or utilize mixed-effects models. To perform mixed-effects models in \pkg{shorts} for split times, one can use `mixed_model_using_splits()` function. To demonstrate this functionality, we load the `split_times` dataset provided in the \pkg{shorts} package:

@@

Additional information about mixed-effects model performed using the `nlme` package [@R-nlme] can be obtained using `summary()` function:

@@

S3 method `coef()` when applied on mixed-model result will return both the fixed and random effects:

```{r}
coef(m2)
```

To create a simple plot of the model, use S3 `plot()` method:

```{r}
plot(m2) + theme_bw(8)
```

@@

variable_names <- colnames(df)

df <- pivot_longer(df, cols = c(-1, -3)) %>%
  mutate(name = factor(name, levels = variable_names))

@@

Let's consider the same example of an athlete with MSS equal to 9 $ms^{-1}$, TAU equal to 1.3, and MAC equal to 6.92 $ms^{-2}$ performing 40m sprint with velocity estimated using radar run (in this case with 1 Hz sampling rate).

@@

```{r}
plot(m3) + theme_bw(8)
```

@@

Mixed-effects model using radar data is done using `mixed_model_using_radar()` function. To perform mixed model, let's load data that comes with \pkg{shorts} package.

@@

```{r}
plot(m4) + theme_bw(8)
```

## Force-Velocity profile

To create *Force-Velocity Profile* (FVP) using single athlete estimated sprint model parameters (i.e., TAU and MSS), you can use `get_FV_profile()` function. When estimating FVP, athlete body mass (kg) can be set using `bodymass` parameter, while the air resistance parameters can be set using `"..."`, which are forwarded to the `get_air_resistance()` function. Details of the FVP method implemented in the \pkg{shorts} package, as well as the interpretation from a sprint training perspective, are covered elsewhere [@haugenPowerForceVelocityProfilingSprinting2020; @morinInterpretingPowerForceVelocityProfiles2016; @morinSimpleMethodComputing2019; @samozinoSimpleMethodMeasuring2016].

```{r}
# To create Force-Velocity Profile
fvp <- get_FV_profile(
  MSS = m1$parameters$MSS,
  TAU = m1$parameters$TAU,
  bodymass = 80,
  # Additional parameters forwarded to get_air_resistance
  # Otherwise, defaults are used
  bodyheight = 1.85,
  barometric_pressure = 780,
  air_temperature = 20,
  wind_velocity = 0.5
)

fvp
```

To plot FVP kinematics and kinetics (which are exactly the same as generated by the `predict_kinematics()` function), use S3 `plot()` function. By default, FVP estimated kinetics are plotted against velocity (on x-axis).

```{r}
plot(fvp) + theme_bw(8)
```

To plot FVP estimated kinetics against time, use `type = "time"` parameter:

```{r}
plot(fvp, "time") + theme_bw(8)
```

@@

One source of error in the modeled estimation using a radar gun is the time synchronization. In theory, synchronization is ideal when a sprint is initiated at $t=0$ (i.e., $v(t=0) = 0$). In practice, this is often not the case. Let's use our athlete and add and deduct 0.5 s to simulate an error in synchronization and its effect on estimated MSS and TAU.

@@

As can be seen from the example, all estimated parameters are affected by an error in synchronization of time with velocity (with MSS being the least affected in this example). The potential solution incorporated into the \pkg{shorts} package involves estimation of the *time correction* parameter using the following equation:

@@

This model is incorporated in the `model_using_radar_with_time_correction()` function:

@@

When using `predict_XXX()` family of functions, one can provide estimated time correction to get predictions at original time scale.

@@

When it comes to mixed-model approach, time correction can be modeled as a fixed effect or random effect using the `mixed_model_using_radar_with_time_correction()` function.

@@

Let's imagine we have two twin brothers with same short sprint characteristics: MSS equal to 9 $ms^{-1}$, TAU equal to 1.3, and MAC equal to 6.92 $ms^{-2}$. Let's call them John and Jack. They both perform 40m sprint using timing gates set at 5, 10, 20, 30, and 40 m. The initial timing gate at the start (i.e., $d=0$ m) serves to activate the timing system (i.e., when they cross the beam).

John represents the *theoretical model*, in which we assume that the initial force production and the timing initiation are perfectly synchronized. Jack, on the other hand, represents a *practical model*, and decides to move slightly behind the initial timing gate (i.e. for 0.5 m) and use body rocking to initiate the sprint start. In other words, Jack is using a *flying start*, a common scenario when testing field sports athletes. Let's see how their sprint outcomes differ.

@@

sprint_parameters <- rbind(
  coef(john_profile),
  coef(jack_profile)
)

@@

Below is a simulation sprint with 5, 10, 20, 30, 40, and 50 m splits, with MSS and MAC varying from 6 to 9 ($ms^{-1}$ and $ms^{-2}$ respectively), and flying start distance varying from 0 to 1 m.

@@

# Add small noise to allow model fit
set.seed(1667)
rand_noise <- rnorm(nrow(sim_df), 0, 10^-5)
sim_df$time <- sim_df$time + rand_noise
```

@@

  params <- data.frame(t(coef(model)))

@@

The following figure demonstrates the effect of flying start distance on estimated MSS:

@@

As can be seen from the figure, MSS is underestimated as flying start distance increases. The following image demonstrates the effect of flying start distance on estimated MAC:

@@

Model residuals are also affected by flying start distance. The shape of residuals distribution depends on number and splits utilized (e.g., 10, 20, 30, 40 m versus 5, 15, 30 m), but here we can see the effect of the flying start distance on the model residuals per split distance utilized in our simulation:

@@

ggplot(
  model_df,
  aes(
    y = residuals,
    x = distance,
    color = flying_start_distance,
    group = group)
) +
  theme_bw(8) +
  geom_line(alpha = 0.3) +
  facet_grid(MSS_string ~ MAC_string) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  scale_color_gradientn(colours = terrain.colors(5, rev = FALSE)) +
  xlab("Distance (m)") +
  ylab("Predicted time - observed time (s)") +
  theme(legend.position = "top") + 
  labs(color = "Flying start distance")
```

If we merge individual facets (i.e., combinations of MSS and MAC), we can get simpler figure conveying issues with residuals when there is a flying start:

```{r}
ggplot(
  model_df,
  aes(
    y = residuals,
    x = distance,
    color = flying_start_distance,
    group = group)
) +
  theme_bw(8) +
  geom_line(alpha = 0.3) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  scale_color_gradientn(colours = terrain.colors(5, rev = FALSE)) +
  xlab("Distance (m)") +
  ylab("Predicted time - observed time (s)") +
  theme(legend.position = "top") + 
  labs(color = "Flying start distance")
```

@@

A potential solution is to use a correction factor - the recommendation in the literature is +0.5 s [@haugenSprintMechanicalProperties2020; @haugenSprintMechanicalVariables2019]. Interestingly, the average difference between using timing gates and a block start for 40 m sprint time was 0.27 s [@haugenDifferenceStartImpact2012]. So, while a timing correction factor is warranted to avoid subsequent errors in estimates of kinetic variables (e.g., overestimate power), a correction factor that is too large will have the opposite effect (e.g., underestimate power).

Rather than providing *apriori* time correction from the literature, \pkg{shorts} package provides an estimation of this parameter from the data provided, together with MSS and TAU. Exactly the same method is suggested by @stenrothForcevelocityProfilingIce2020, named *time shift method*, and the estimated parameter named *time shift parameter*. We have named this parameter *time correction* to be in agreement with the parameter introduced in [Problems with time sync with radar gun] section of this paper, as well as the available literature.

When implementing time correction, equation \@ref(eq:time-distance) becomes:

@@

To estimate time correction parameter, we use `model_using_splits_with_time_correction()` function. Here is how we can estimate Jack parameters using either provided time correction (e.g., +0.3 and +0.5 s) or estimated time correction:

@@

jack_parameters <- rbind(
  coef(john_profile),
  coef(jack_profile),
  coef(jack_profile_fixed_time_short),
  coef(jack_profile_fixed_time_long),
  coef(jack_profile_time_estimated)
)

@@

In Jack's case, both +0.3 s fixed time correction and time correction estimation yield parameters closer to John's (i.e. true parameters).

Another model definition, which is a novel approach implemented in the \pkg{shorts} package, is to utilize *distance correction*, besides time correction. Thus, equation \@ref(eq:time-distance) becomes:

@@

This model is implemented in `model_using_splits_with_corrections()` function. Below are the model estimates:

@@

jack_parameters <- rbind(
  coef(john_profile),
  coef(jack_profile),
  coef(jack_profile_fixed_time_short),
  coef(jack_profile_fixed_time_long),
  coef(jack_profile_time_estimated),
  coef(jack_profile_distance_correction)
)

@@

As can be seen from the results, adding distance correction results in correctly estimating Jack's sprint parameters. There are a few issues with this model definition. Besides being novel and still not validated with actual data, distance correction model has four parameters to estimate, which implies that at least five sprint splits are needed. This imposes practical limitations, since acquiring six timing gate (one for the start and five for splits) might be practically troublesome. One strategy that is sometimes implemented is adding zeros to the sample (i.e., $t=0$ and $d=0$), which increase the number of observations. Unfortunately, this strategy should not be implemented, as explained later in the [Should we add zero to the sample?] section of this paper.

@@

  params <- rbind(
    data.frame(
      model = "No correction",
      t(coef(no_correction))
    ),
    data.frame(
      model = "Fixed correction +0.3s",
      t(coef(fixed_correction_short))
    ),
    data.frame(
      model = "Fixed correction +0.5s",
      t(coef(fixed_correction_long))
    ),
    data.frame(
      model = "Time correction",
      t(coef(time_correction))
    ),
    data.frame(
      model = "Time and distance correction",
      t(coef(time_dist_correction))
    )
  )

@@

ggplot(
  model_df,
  aes(
    y = residuals,
    x = distance,
    color = flying_start_distance,
    group = group)
) +
  theme_bw(8) +
  geom_line(alpha = 0.3) +
  facet_wrap(~model) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  scale_color_gradientn(colours = terrain.colors(5, rev = FALSE)) +
  xlab("Distance (m)") +
  ylab("Predicted time - observed time (s)") +
  theme(legend.position = "top") + 
  labs(color = "Flying start distance")
```

@@

Time correction and time and distance corrections are also implemented in the mixed-models using `mixed_model_using_splits_with_time_correction()` and `mixed_model_using_splits_with_corrections()`. We will showcase their use at the end of this paper.

@@

Starting behind the initial timing gate represent only one issue (i.e., flying start). In this section, we simulate one more issue to check the sensitivity of the presented models to other (less common) perturbations when performing field testing.

One issue that might happen with timing gates is triggering the timing system before the sprint is initiated (e.g., by cutting the beam with an arm swing prematurely). This is very similar to the situation when timing starts on a signal (i.e., gun during 100 m sprint race) and there is *reaction time* (RT) involved. Both of these scenarios represent *time lag* that is added to the split times. Below we simulate the effect of this time lag on model estimates and predictions.
